---
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup1, include=FALSE}
knitr::opts_chunk$set(fig.path = "man/figures/README_", warning = FALSE, message = FALSE, error = FALSE, echo = TRUE)
set.seed(1)
library(tidyverse)
```

# princurve
[![Build Status](https://travis-ci.org/dynverse/princurve.svg?branch=master)](https://travis-ci.org/dynverse/princurve)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/dynverse/princurve?branch=master&svg=true)](https://ci.appveyor.com/project/dynverse/princurve)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/princurve)](https://cran.r-project.org/package=princurve)
[![Coverage Status](https://codecov.io/gh/dynverse/princurve/branch/master/graph/badge.svg)](https://codecov.io/gh/dynverse/princurve?branch=master)

Fitting a principal curve to a data matrix in arbitrary dimensions.

## Example
We generate some example data:
```{r example_data}
t <- runif(100, -1, 1)
x <- cbind(t, t ^ 2) + rnorm(200, sd = 0.05)
colnames(x) <- c("dim1", "dim2")

plot(x)
```

A principal curve can be fit to the data as follows:
```{r example_plot}
library(princurve)
fit <- principal_curve(x)
plot(fit); whiskers(x, fit$s, col = "gray")
```

Check out `?principal_curve` for more information on the specific outputs of `principal_curve()`. 
An overview of the principal curve algorithm is given in the [algorithm](vignettes/algorithm.md) vignette.

## Benchmarks

Using the experimental `approx_points` parameter, princurve 2.1.0 offers major performance improvements for large datasets.

```{r compare, fig.height=8, fig.width=8}
data("benchmarks", package = "princurve")
ggplot(benchmarks, aes(num_points, median / 1000)) +
  geom_point() +
  geom_line() +
  facet_wrap(~expr, ncol = 1, scales = "free") +
  theme_bw() +
  labs(x = "Number of rows in dataset", y = "Time (s)") +
  scale_colour_brewer(palette = "Set1")
```

Read more about this feature in the [benchmarks](vignettes/benchmarks.md) vignette.

<!-- ## Latest changes -->
<!-- This section gets automatically generated from inst/NEWS.md, and also generates inst/NEWS -->

```{r news, results='asis', echo=FALSE}
news_md <- readr::read_lines("inst/NEWS.md")

# creating NEWS for package
news_normal <- news_md %>% 
  str_replace_all("^# princurve", "princurve") %>% 
  str_replace_all("\\[[^\\]]*\\]\\(([^\\)]*)\\)", "\\1")
readr::write_lines(news_normal, "inst/NEWS")

# creating text for readme
ix <- which(str_detect(news_md, "^# princurve"))

cat(str_replace(news_md[[ix[[1]]]], "^# ", "## Latest changes in "))

cat(news_md[seq(ix[[1]]+1, ix[[2]]-1)], sep = "\n")

cat(str_replace(news_md[[ix[[2]]]], "^# ", "## Latest changes in "))

cat(news_md[seq(ix[[2]]+1, ix[[3]]-1)], sep = "\n")
```

Check [NEWS.md](inst/NEWS.md) for a full list of changes.

## References

Hastie, T. and Stuetzle, W., [Principal Curves](https://www.jstor.org/stable/2289936), JASA, Vol. 84, No. 406 (Jun., 1989), pp. 502-516, DOI: [10.2307/2289936](http://doi.org/10.2307/2289936) ([PDF](https://web.stanford.edu/~hastie/Papers/principalcurves.pdf))
