---
title: "Benchmarking princurve"
author: "Robrecht Cannoodt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking princurve}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

<!-- github markdown using
rmarkdown::render("vignettes/benchmarks.Rmd", output_format = "github_document")
-->

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
set.seed(1)
```

```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(purrr)
library(ggplot2)
library(princurve)
data("benchmarks", package = "princurve")
```

princurve 2.1 contains major optimisations if the `approx_points` parameter is used. 
This is showcased on a toy example, where the number of points was varied between $10^2$ and $10^6$.

```{r benchmark, eval=FALSE}
# warning: running this will take ± 2 hours
# princurvelegacy can be installed using devtools::install_github("dynverse/princurve@legacy")
num_points <- round(10^seq(log10(100), log10(100000), length.out = 25))
lambda <- rnorm(max(num_points), 0, .2)
x <- cbind(lambda, lambda^2) + rnorm(length(lambda) * 2, 0, .02)

benchmarks <- map_df(
  rev(num_points),
  function(np) {
    xsel <- x[seq_len(np),]
    microbenchmark::microbenchmark(
      "princurve1.1" = princurvelegacy::principal.curve(xsel), 
      "princurve2.1" = princurve::principal_curve(xsel, approx_points = 100),
      times = 10L,
      unit = "ms"
    ) %>%
      summary() %>%
      as_data_frame() %>%
      mutate(num_points = np)
  }
)
```

We can see princurve 2.1 scales linearly w.r.t. the number of rows in 
the dataset, whereas princurve 1.1 scales quadratically.
This is due to the addition of the approximation step added in between 
the smoothing and the projection steps (explained in more detail 
in the [algorithm](algorithm.md) vignette).
```{r compare, fig.width=8, fig.height=8}
ggplot(benchmarks, aes(num_points, median / 1000)) +
  geom_point() +
  geom_line() +
  facet_wrap(~expr, ncol = 1, scales = "free") +
  theme_bw() +
  labs(x = "Number of rows in dataset", y = "Time (s)") +
  scale_colour_brewer(palette = "Set1")
```
