---
title: "Introduction to the princurve package"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: false
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to the princurve package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- render using
rmarkdown::render(input = "vignettes/intro.Rmd")
-->

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(comment = "#>", echo = FALSE, fig.width = 8, fig.height = 6)
set.seed(1)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(princurve)

library(magick)

ggif_list <- function(list, .width = 8, .height = 6, .dpi = 120, .fps = 1, ...) {
  dir <- tempfile("gif_files")
  dir.create(dir)
  on.exit(unlink(dir))
  
  img <- lapply(
    seq_along(list),
    function(i) {
      filename <- paste0(dir, "/image-", i, ".png")
      ggsave(filename, list[[i]], width = .width, height = .height, dpi = .dpi)
      image_read(filename)
    }
  )
  
  image_animate(do.call(c, img), fps = .fps, dispose = "none")
}

ggif_lapply <- function(X, FUN, .width = 8, .height = 6, .dpi = 120, .fps = 1, ...) {
  list <- lapply(X, FUN)
  ggif_list(list, .width = .width, .height = .height, .dpi = .dpi, .fps = .fps, ...)
}
```

A principal curve is a smooth curve passing through the middle of a multidimensional dataset. 
This package is an R/C++ reimplementation of the S/Fortran code provided by Trevor Hastie, 
with multiple performance tweaks.

## Example
Deriving a principal curve is an iterative process. This is what it looks like for 
a two-dimensional toy dataset:

```{r dataset}
set.seed(1)
z <- sort(runif(100, -1.4 * pi, .4 * pi))
s <- data_frame(
  x = cos(z) * 1.5,
  y = sin(z)
)
x <- s %>% 
  sample_frac(1) %>% 
  mutate(
    x = x + rnorm(length(x), 0, .05),
    y = y + rnorm(length(x), 0, .05)
  )
```

```{r iterative}
ggif_lapply(seq(0, 10), function(it) {
  fit <- principal_curve(as.matrix(x), maxit = it)
  
  curve <- 
    as_data_frame(fit$s) %>% 
    mutate(lambda = fit$lambda, it = it) %>% 
    slice(fit$ord) %>% 
    mutate(pos = seq_len(n()))
  
  ggplot() +
    geom_point(aes(x, y), x, colour = "darkgray") +
    geom_path(aes(x, y), curve) +
    theme_bw() +
    coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
    labs(title = paste0("Iteration ", it)) 
})
```


## Algorithm
Pseudocode for the princurves algorithm is shown below.
The individual steps will be explained in more detail in the
following subsections.

```r
# initialisation
s = principal_component(x)
x_proj = project(x, s)
lambda = arc_length(x_proj)

# iterative process
for (it = 1..max_iter) {
  s = smooth(lambda, x)
  s' = approximate(lambda, s, num_points = 100)
  x_proj = project(x, s')
  lambda = arc_length(x_proj)
}
```


### Initialisation
The principal curve `s` is initialised (at iteration 0) by
calculating the principal component. All points in `x` are projected
orthogonally onto `s`, and the arc-length `lambda` of each projection w.r.t.
to the start of the curve is calculated.

```{r initialisation}
fit0 <- principal_curve(as.matrix(x), maxit = 0)

steps <- c(
  "Step 0a: Initialise curve with principal component", 
  "Step 0b: Orthogonally project points to curve",
  "Step 0c: Calculate arc-length of projections w.r.t. the origin of the curve"
)

g0 <- ggplot() +
  geom_point(aes(x, y), x) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = paste0(c("", "", ""), steps, collapse = "\n"))

g1 <- ggplot() +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_path(aes(x, y), as_data_frame(fit0$s[fit0$ord, ])) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = paste0(c("> ", "", ""), steps, collapse = "\n"))

g2 <- ggplot() +
  geom_segment(aes(x = x$x, xend = fit0$s[,1], y = x$y, yend = fit0$s[,2]), linetype = "dashed") +
  geom_path(aes(x, y), as_data_frame(fit0$s[fit0$ord, ]), colour = "darkgray") +
  geom_point(aes(x, y), x, colour = "darkgray") +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = paste0(c("", "> ", ""), steps, collapse = "\n"))

g3 <- ggplot() +
  geom_segment(aes(x = x$x, xend = fit0$s[,1], y = x$y, yend = fit0$s[,2]), linetype = "dashed", colour = "lightgray") +
  geom_path(aes(x, y), as_data_frame(fit0$s[fit0$ord, ]), colour = "darkgray") +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_point(aes(x, y), as_data_frame(fit0$s)) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = paste0(c("", "", "> "), steps, collapse = "\n"))

ggif_list(list(g0, g1, g2, g3))
```

### Iteration 1
Each iteration consists of three steps: smoothing, approximation, and projection.

#### Smoothing: calculate new curve
During the smoothing step, a new curve is computed by smoothing each 
dimension in `x` w.r.t. the arc-length `lambda` calculated for the previous curve.

```{r smooth}
fit1 <- principal_curve(as.matrix(x), maxit = 1)

xdf <- x %>% 
  mutate(lambda = fit0$lambda) %>%
  arrange(lambda) %>% 
  gather(dimension, value, -lambda) %>% 
  group_by(dimension) %>% 
  mutate(smooth = smooth.spline(lambda + runif(length(lambda), 0, .001), value, df = 5)$y) %>% 
  ungroup()

ggplot(xdf) +
  geom_point(aes(lambda, value)) +
  geom_line(aes(lambda, smooth)) +
  facet_wrap(~dimension, scales = "free") +
  theme_bw()
```

```{r beforeafter}
g0 <- ggplot() +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_path(aes(x, y), as_data_frame(fit0$s[fit0$ord, ])) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = "Before")

g1 <- ggplot() +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_path(aes(x, y), as_data_frame(fit1$s[fit1$ord, ])) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = "After")

ggif_list(list(g0, g1))
```

#### Approximation: simplify curve
In the next step (projection), each of the `n` points in `x` will get compared to each segment in the curve `s`. 
After the smoothing step, the curve consists of `n` points. Thus, the projection step would have a quadratic complexity.
In order to make this step to behave more linearly, the `approx_points = 100` parameter can be used to 
first approximate by a curve with 100 points.

```{r approx}
xout <- seq(min(xdf$lambda), max(xdf$lambda), length.out = 100)
xadf <- xdf %>% 
  group_by(dimension) %>% 
  do({
    data_frame(lambda = xout, dimension = .$dimension[[1]], smooth = stats::approx(.$lambda, .$smooth, xout)$y)
  }) %>% 
  ungroup() %>% 
  spread(dimension, smooth)

g0 <- ggplot() +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_path(aes(x, y), as_data_frame(fit1$s[fit1$ord, ])) +
  geom_point(aes(x, y), as_data_frame(fit1$s)) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = "Before")

g1 <- ggplot() +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_path(aes(x, y), xadf) +
  geom_point(aes(x, y), xadf) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = "After")

ggif_list(list(g0, g1))
```

#### Projection: calculate new lambda
The projection step is same as before; all the points are orthogonally 
projected onto the new curve, and the arc-length `lambda` is recalculated 
for the new projections.

```{r relambda}
ggplot() +
  geom_segment(aes(x = x$x, xend = fit1$s[,1], y = x$y, yend = fit1$s[,2]), linetype = "dashed", colour = "lightgray") +
  geom_path(aes(x, y), as_data_frame(fit1$s[fit1$ord, ]), colour = "darkgray") +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_point(aes(x, y), as_data_frame(fit1$s)) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y")
```

This process is repeated until convergence or until a predefined number of iterations has passed.

### Iteration 2
For clarity's sake, the smoothing and projection steps are also shown for iteration 2.

#### Smoothing: calculate new curve
During the smoothing step, a new curve is computed by smoothing each 
dimension in `x` w.r.t. the arc-length `lambda` calculated for the previous curve.

```{r smooth2}
fit2 <- principal_curve(as.matrix(x), maxit = 2)

xdf2 <- x %>% 
  mutate(lambda = fit1$lambda) %>%
  arrange(lambda) %>% 
  gather(dimension, value, -lambda) %>% 
  group_by(dimension) %>% 
  mutate(smooth = smooth.spline(lambda + runif(length(lambda), 0, .001), value, df = 5)$y) %>% 
  ungroup()

ggplot(xdf2) +
  geom_point(aes(lambda, value)) +
  geom_line(aes(lambda, smooth)) +
  facet_wrap(~dimension, scales = "free") +
  theme_bw()
```

```{r beforeafter2}
g0 <- ggplot() +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_path(aes(x, y), as_data_frame(fit1$s[fit1$ord, ])) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = "Before")

g1 <- ggplot() +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_path(aes(x, y), as_data_frame(fit2$s[fit2$ord, ])) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = "After")

ggif_list(list(g0, g1))
```

#### Approximation: simplify curve
The curve is simplified in order to make the projection step easier.

```{r approx2}
xout2 <- seq(min(xdf2$lambda), max(xdf2$lambda), length.out = 100)
xadf2 <- xdf2 %>% 
  group_by(dimension) %>% 
  do({
    data_frame(lambda = xout2, dimension = .$dimension[[1]], smooth = stats::approx(.$lambda, .$smooth, xout2)$y)
  }) %>% 
  ungroup() %>% 
  spread(dimension, smooth)

g0 <- ggplot() +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_path(aes(x, y), as_data_frame(fit2$s[fit2$ord, ])) +
  geom_point(aes(x, y), as_data_frame(fit2$s)) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = "Before")

g1 <- ggplot() +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_path(aes(x, y), xadf2) +
  geom_point(aes(x, y), xadf2) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y", title = "After")

ggif_list(list(g0, g1))
```

#### Projection: calculate new lambda
All the points are orthogonally 
projected onto the new curve, and the arc-length `lambda` is recalculated 
for the new projections.

```{r relambda2}
ggplot() +
  geom_segment(aes(x = x$x, xend = fit2$s[,1], y = x$y, yend = fit2$s[,2]), linetype = "dashed", colour = "lightgray") +
  geom_path(aes(x, y), as_data_frame(fit2$s[fit2$ord, ]), colour = "darkgray") +
  geom_point(aes(x, y), x, colour = "darkgray") +
  geom_point(aes(x, y), as_data_frame(fit2$s)) +
  theme_bw() +
  coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-1.1, 1.1)) +
  labs(x = "x", y = "y")
```


## Timing comparison
princurve 2.1 contains some major optimisations, if the `approx_points` parameter is used. 
This is showcased on a toy example, where the number of points was varied between $10^2$ and $10^6$.

We can see princurve 2.1 scales quasi linearly w.r.t. the number of rows in 
the dataset, whereas princurve 1.1 scales quadratically.
This is due to the addition of the approximation step added in between 
the smoothing and the projection steps.

```{r compare, fig.width=8, fig.height=8}
data("benchmarks", package = "princurve")

ggplot(benchmarks, aes(num_points, median / 1000)) +
  geom_point() +
  geom_line() +
  facet_wrap(~expr, ncol = 1, scales = "free") +
  theme_bw() +
  labs(x = "Number of rows in dataset", y = "Time (s)") +
  scale_colour_brewer(palette = "Set1")
```
